---
title: "p8105_hw2_jz3902"
author: "Jinghan Zhao"
date: "2024-09-27"
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
```

## Problem 1

### Import and clean the data

The original dataset contains variables including: division, line, station name, station latitude / longitude, routes served, entrance type, entry, vending, staff and staff hours, ADA compliance and notes, free crossover, north south / east west street, corner, entrance latitude / longitude, station location and entrance location.

Data cleaning steps: clean variable names -> select variables in interest -> convert the `entry` variable from character to a logical variable by `case_match()`.

```{r data_cleaning, message = FALSE}
transit_df = 
  read_csv("HW2_data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% 
  janitor::clean_names() %>% 
  select(line:entry, vending, ada) %>% 
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

```

The resulting dataset has `r nrow(transit_df)` rows and `r ncol(transit_df)` columns, which contains variables including: line, station name, station latitude / longitude, routes served, entrance type, entry, vending and ADA compliance.


### Distinct stations

Distinct stations are defined by line and station name.

```{r distinct_stations}
transit_df %>% 
  distinct(line, station_name) %>% 
  nrow()
```

There are 465 distinct stations.


### ADA compliant

```{r ada_compliant}
transit_df %>%
  filter(ada == TRUE) %>% 
  nrow()
```

There are 468 ADA compliant stations.


### Vending

Calculate the numerator (allow entrance + no vending) and denominator (no vending) respectively, then calculate the proportion.

```{r vending}
no_vending = 
  transit_df %>%
  filter(vending == "NO") %>% 
  nrow()

entry_no_vending = 
  transit_df %>%
  filter(vending == "NO", entry == TRUE) %>% 
  nrow()

propo_vending = entry_no_vending / no_vending
```

There are `r propo_vending * 100`% of stations without vending allow entrance.


### Display routes

The original variable type of `route1:route7` are character, yet `route8:route11` are numeric, so they cannot combine together. Therefore, I convert `route8:route11` to character first, then do `pivot_longer()`.

```{r routes}
transit_tidy_df =
  transit_df %>% 
  mutate(
    across(route8:route11, as.character)
  ) %>% 
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE)
```

Answering questions:

```{r routes_question}
transit_tidy_df %>% 
  filter(route_name == "A") %>% 
  distinct(line, station_name) %>% 
  nrow()

transit_tidy_df %>% 
  filter(route_name == "A", ada == TRUE) %>% 
  nrow()
```

There are 60 distinct stations serve the A train.

There are 107 ADA compliant stations that serve the A train.




